---
title: "Cabingan_hclust"
author: "Stephanie Cabingan"
date: "2026-06-10"
output: html_document
---

A. 
```{r}
# install.packages("cluster")
library(cluster)
```

Read Data
```{r}
df <- read.csv("morphology_data.csv")
```

Exclude sex and population column
```{r}
clmph_data <- df[,3:19]
clmph_data
```

Cleaning
```{r}
good <- complete.cases(clmph_data)
head(clmph_data[good, ])
morpho_data <- clmph_data[good,]
```

Standardization
```{r}
morpho_scaled <- scale(morpho_data)
```

Euclidean Distance
```{r}
d <- dist(morpho_scaled, method = "euclidean")
```

Single Linkage
```{r, fig.width = 20, fig.height = 6}
mp_single <- hclust(d, method = "single")
rownames(morpho_scaled) <- df$Pop

plot(mp_single,
     main = "Single Linkage",
     labels = rownames(morpho_scaled),
     cex = 0.7)
```

Ward's Linkage
```{r, fig.width = 20, fig.height = 6}
mp_ward <- hclust(d, method = "ward.D2")

plot(mp_ward,
     main = "Ward D2",
     labels = rownames(morpho_scaled),
     cex = 0.7)
```

Single Linkage Loop
```{r, fig.width = 8, fig.height = 6}
mpsingle_loop <- numeric(10)

for(k in 2:10){
  mp_clusters <- cutree(mp_single, k)
  mp_sil <- silhouette(mp_clusters, d)
  mpsingle_loop[k] <- mean(mp_sil[, 3])
}

plot(
    mpsingle_loop,
    type = "b",
    xlab = "Cluster No.",
    ylab = "Silhouette Width",
    main = "Silhouette Scores for Single Linkage Method"
)
```

Ward Linkage Loop
```{r, fig.width = 8, fig.height = 6}
mpw_loop <- numeric(10)

for(k in 2:10){
  mpw_clusters <- cutree(mp_ward, k)
  mpw_sil <- silhouette(mpw_clusters, d)
  mpw_loop[k] <- mean(mpw_sil[, 3])
}

plot(
  mpw_loop,
  type = "b",
  xlab = "Cluster No.",
  ylab = "Silhouette Width",
  main = "Silhouette Scores for Ward Method"
)
```

Single Linkage Loop Dendrogram
```{r, fig.width = 20, fig.height = 6}
opk_single <- which.max(mpsingle_loop) + 1
rownames(morpho_scaled) <- df$Pop

plot(
  mp_single,
  main = "Single Linkage Loop with K = 2",
  labels = rownames(morpho_scaled),
  cex = 0.7
)

rect.hclust(
  mp_single,
  k = opk_single,
  border = c("lightblue", "pink")
)
```

Ward Linkage Loop Dendrogram
```{r, fig.width = 20, fig.height = 6}
opk_ward <- which.max(mpw_loop) + 1
rownames(morpho_scaled) <- df$Pop

plot(
  mp_ward,
  main = "Ward Linkage Loop with K = 2",
  labels = rownames(morpho_scaled),
  cex = 0.7
)

rect.hclust(
  mp_ward,
  k = opk_ward,
  border = c("green", "yellow")
)
```

B.

1. For Activity 7, k was determined through the elbow curve and had a value of 3, but this time it was determined through the silhouette plot and results in a value of 2.

2. No the clusters were not consistent. I think because all the different functions do different things so it's important to maybe go through all the methods to make sure. 