---
title: "Activity 3 R Markdown"
author: "Stephanie Cabingan"
date: "2026-06-01"
output: html_document
---

#Read Packages
```{r packages, include=FALSE}
library(rmarkdown)
library(knitr)
```

# Read Data
```{r}
df <- read.csv("herpetozoa-033-095-s001.csv")
```

# Cleaning and Data Structure
```{r}
good <- complete.cases(df)
kable(head(df[good, ]))
herp_data <- df[good, ]

# Data Structure
kable(class(herp_data))
str(herp_data)
```

# Elevation Data Subset
```{r}
elev_data <- herp_data[,5]
elev_data
```

# Generate Mean, SD, and Range for Elevetion Data  Subset
```{r}
mean_elev <- mean(elev_data)
sd_elev <- sd(elev_data)
rng_elev <- range(elev_data)

mean_elev
sd_elev
rng_elev
```

# Species Subset
```{r}
sp_data <- herp_data[,7:33]
```

# Typical Loop
```{r loops}
col_loop <- numeric(ncol(sp_data))

for (i in 1:ncol(sp_data)){
  col_loop[i] <- sum(sp_data[, i])
}

names(col_loop) <- colnames(sp_data)
kable(col_loop, digits = 2)
```

# Compact Loop
```{r}
sp_sapply <- sapply(sp_data,sum)
kable(sp_sapply, digits = 2)
```

# Minimum and Maximum Values
```{r minmax}
max_spnum <- max(col_loop)
max_sp <- names(col_loop[col_loop == max_spnum])

min_spnum <- min(col_loop)
min_sp <- names(col_loop[col_loop == min_spnum])

max_sp
max_spnum
min_sp
min_spnum
```

# Combine Most Observed Species and Least Observed Species with Habitat Types
```{r}
habitat_types <- herp_data[,4]
spht_data <- data.frame(habitat_types, sp_data[max_sp], sp_data[min_sp])
kable(spht_data, digits = 2)
```

# Mean Count of each species per habitat
```{r}
mean_spht <- aggregate(cbind(Phlo, Babu, Geat, Hety, Bosc, Drru, Apbo) ~ habitat_types, data = spht_data, FUN = mean)
kable(mean_spht, digits = 2)
```

# Combine Species Total Count, Elevation, and Habitat Type
```{r}
species_total <- herp_data[,6]
elev_info <- herp_data[,5]
plot_data <- data.frame(species_total, elev_data, habitat_types)
kable(plot_data, digits = 2)
```

# Scatterplot
```{r, fig.widht = 10, fig.height = 4}
plot_color <- as.numeric(as.factor(habitat_types))
plot_graph <- plot(elev_info, species_total, col = plot_color, main = "Species Totals per Elevation", xlab = "Elevation", ylab = "Species Totals", pch = 16)
plot_graph
```

# Bar Graph
```{r}
bar_sptotal <- tapply(species_total, habitat_types, sum)
barplot(bar_sptotal, xlab = "Habitat Types", ylab = "Species totals", names.arg = c("ASG", "CVT", "ESG", "OGF"))
```

# Histogram
``` {r}
hist(species_total, main = "Total Species Counts", xlab = "Total Species Counts", ylab = "Frequency")
```

## Analysis
A negative correlation can be seen among elevation and total species. This indicates that greater species richness can be found in lower levels while becoming less and less as elevation increases. What can also be gathered is that habitat types at lower levels tend to hold greater species numbers. This indicates that a certain habitat type could be more favorable in containing a greater number of species.