---
title: "CABINGAN_Act4"
author: "Stephanie Cabingan"
date: "2026-06-01"
output: html_document
---

# Install and load Tidyverse, R Markdown, and Knitr
```{r}
library(tidyverse)
library(rmarkdown)
library(knitr)
```

# Read the csv file
```{r}
df <- read.csv("habitat_types.csv")
```


# Show data structure
```{r}
df |> glimpse()
```

# Subset for Transect, Htypes, SmallTrees, FloweringTrees
```{r}
sub_set <- select(df, Transect, Htypes, SmallTrees, FloweringTrees)
kable(sub_set, digit = 2)
```

# CVT Only in descending order
```{r}
ht_st <- filter(sub_set, Htypes == "CVT")
sort_htst <- arrange(ht_st, desc(SmallTrees))
sort_htst |> slice(c(1,1))
```


# Mean of flowering trees
```{r}
sub_set |> 
  group_by(Htypes) |>
  summarise(mean_flowering = mean(FloweringTrees)) -> mean_ft
  mean_ft |> slice(c(1,3))
```

# New Column "HerbGroundDiff"
```{r}
sub_set |> 
  mutate(HerbGroundDiff = df$HerbsCover - df$GroundCover) -> sel_df_mod
  sel_df_mod  

sel_df_mod |>
  filter(HerbGroundDiff == 0) -> tf_diff
  tf_diff

tf_diff |>
  select(Transect, HerbGroundDiff)|>
  head(3)
```

# Wider sel_df_mod
```{r}
sel_df_mod |>
  pivot_wider(names_from = Htypes, values_from = c(SmallTrees, FloweringTrees, HerbGroundDiff)) -> sdm_wide
sdm_wide
```

# Analysis
Comparing to base R, using tidyverse feels better for me, and it has a familiar feel to Python (that I cannot really explain.) It feels easier to understand, wherein using base R, I struggle to track what exactly a line of code is doing, with tidyverse, everything seems easier to figure out.