---
title: "Cabingan_hclust"
author: "Stephanie Cabingan"
date: "2026-06-17"
output: html_document
---

Read packages
```{r}
library(PCAtest)
library(rmarkdown)
library(knitr)
```

A. 

Read data
```{r}
df <- read.csv("Meneses_et.al_2026_Platymantis.csv")

```

Subset Species and Morphological Data
```{r}
species_data <- df[,-c(2:9)]
kable(species_data)
```

Data Structure and Class
```{r}
class(species_data)
str(species_data)
```

Cleaning data
```{r}
good <- complete.cases(species_data)
head(species_data[good, ])
spmo_data <- species_data[good,]
```

Scaling data
```{r}
numeric_data <- sapply(spmo_data, is.numeric)
spmo_scaled <- scale(spmo_data[,numeric_data])
kable(spmo_scaled)
```

Pairwise correlation
```{r}
spmo_corr <- cor(spmo_scaled)
kable(round(spmo_corr, 2))
```

Eigenvalues
```{r}
eig <- eigen(spmo_corr)
eig
```

Proportion of variance
```{r}
spmo_var <- eig$values / sum(eig$values)
kable(spmo_var)
```

Eigenvalues to percentages of total variance
```{r}
spmo_pervar <- (eig$values / sum(eig$values)) * 100
```

Comptute PC scores
```{r}
eig_vec <- eig$vectors
scores <- spmo_scaled %*% eig_vec
kable(head(scores))
```

Plot
```{r}
spmo_pca <- prcomp(spmo_scaled)
species <- as.factor(spmo_data$Species)
colors <- c("black", "pink3")

par(mfrow = c(1,2))

plot(
  spmo_pca,
  type = "barplot",
  main = "Scree Plot"
)

plot(
  spmo_pca$x[,1],
  spmo_pca$x[,2],
  col = colors,
  pch = c(19,17),
  cex = 0.6,
  xlab = "PC1",
  ylab = "PC2",
  main = "PCA Biplot"
)

legend("topright",
       legend = c("PC1", "PC2"),
       pch = c(19, 17),
       col = colors
)

```

B.

PCATest package
```{r}
pcatest <- PCAtest(spmo_scaled, 100, 100, 0.05, varcorr = FALSE, counter = FALSE, plot = TRUE)
```

What is the difference between the two methods (broken-stick and PCAtest)
- Broken-stick would depend on eigenvalue vectors while PCAtest randomizes the dataset and runs PCA for every randomization.

Broken-stick Barplot
```{r}
broken_stick <- function(p){
  sapply(1:p, function(k){
    sum(1 / (k:p)) / p
  })
}

bs <- broken_stick(ncol(spmo_scaled))

comparison <- data.frame(
  PC = paste0("PC"),
  Observed = spmo_pervar,
  BrokenStick = bs
)

comparison

plot(
    spmo_pervar,
     type = "b",
     pch = 19,
    col = "black",
     ylim = c(0, max(spmo_pervar)),
     xlab = "Principal Component",
     ylab = "Proportion of Variance",
     main = "Broken-Stick Analysis"
)

lines(bs,
      type = "b",
      pch = 17,
      col = "pink3"
)

legend("topright",
       legend = c("Observed", "Broken Stick"),
       pch = c(19, 17),
       col = c("black", "pink3")
)
```